# The Normal Distribution {#norm_dist}

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


```{r}
library(moments)
```

This chapter supplements the content in Lecture 3.

## A Simple Demonstration of the Normal Distribution

First, let's consider the 'empirical distribution', which is the pattern your data set makes.

Let's consider UK births in 2020 for our example, since they're a subject close to my heart (my first child was born in 2020).

Looking online, one can pull down the distribution of UK birthweights in 2020, available from the UK ONS website, and which are reported in 500g 'bins'.

Source: https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/bulletins/birthsummarytablesenglandandwales/2020#births-data

We can use these numbers to create a chart

(please note, the code I use for the demonstrations in this lesson is mostly modified from David Spiegelhalter's originals, available on his Github page, with some of my own bits and bobs and new data...)

Spiegelhalter's Github can be found here: https://github.com/dspiegel29/ArtofStatistics

The first thing I want to do is create a data set that accurately reflects the ONS distribution of UK birthweights in 2020. To do that, I use the following code, which in this instance I will display below, with annotations.


```{r, include=TRUE}
weights=c(1500, 2000, 2500, 3000, 3500, 4000, 4500,5000,5500) #modified categories from 2020 UK Data
mids=weights-250 # Note the DS original is +250, but this does not work for UK data as the numbered is the UPPER bound of the weight category in the ONS stats. So the babies in the '2000' bin weigh 1500-1999, so the midpoint is 1750, not 2250
n=c(5015, 7103, 27554, 99220, 218442, 179088, 53530,6952,635) # numbers in each bin, UK2020
N=sum(n)  # total number of babies
area=N*500  # number * binwidth = total area of histogram

#I think lbw should be sum of groups 1-3 not 1-2 as in the book code. Using 1-3 gives a result for the UK of 6.6% which tallies with the official stat given that I remove the 'no recorded weight' and 'implausible' categories'. I imagine that the DS book data has categories which may be the lower bound not the upper as in the UK data table.

lbw   = sum(n[1:3])   # number with low birth weight (less than 2500)
lbw.percent=100*lbw/N  # % low birth weight
# 6.6% which tallies with Nuffield stat for 2020.

```

Remember though, the above just sets up the numbers required to recreate the empirical distribution, it isn't analysis itself.

Next, I want to summarize the distribution I just created and a few important numbers regarding Low Birthweight that we will need to use for our later discussion.

Displayed below are the total N for the distribution, the number of low birthweight (LBW) babies born, the percentage of LBW babies, the mean, the standard deviation, the skewness, and kurtosis.

```{r}
print(N)
print(lbw)
print(lbw.percent)
```

Next, calculate the mean and standard deviation of the empirical distribution:

```{r}
#calculate mean and sd of population
# could use sheppard's correction

birth.mean=sum(n*mids/N)
birth.sd=sqrt( sum(n*(mids-birth.mean)^2)/N)
```

```{r}
print(birth.mean) # note as a sense check this mean tallies broadly with that reported in the media around that time which is around 7lb 8oz - 3401g. Slight neg skew expected here I guess due to the mids calculation.
print(birth.sd)
```

```{r}
# just for my own info
skewness(weights)
kurtosis(weights)
```

Next, I will calculate the estimated proportion of low birthweight babies using a normal distribution approximation, and see if it tallies OK with the figure that is aken from the distribution of 6.6%

I will also calculate the middle birthweight proportion as a sense check - should be very close to 50%!

```{r}

# per cent less than 2500 from normal approximation
lbw.est = 100 * pnorm(2500,birth.mean, birth.sd)
# 7.2%, not a bad approximation for UK figure which is about 6.6% from nuffield

mbw.est = 100 * pnorm(3345, birth.mean, birth.sd)

```

```{r}
print(lbw.est)
print(mbw.est)
```

For our reference, we will calculate the 25th, 50th, 75th percentiles

```{r}

#25th, 50th and 75th percentiles of population
qnorm(0.25, birth.mean,birth.sd)
qnorm(0.50, birth.mean,birth.sd)
qnorm(0.75, birth.mean, birth.sd)
```

Next I want to calculate the birthweight percentile of Monty, my son.

```{r}
# percentile of baby weighing MONTY
xw = 3912
pnorm(xw, birth.mean,birth.sd)
#83rd percentile. Fairly large. Tallies with the numbers on record.
```

Now, at last, we can do some plots!

First let's set up the histogram and overlay a normal distribution on to it.

The weight of Monty is overlaid as a red line. We can estimate if this is 'unusual' or not by how far away it is from the mean of the distribution.

But, what actually is a normal distribution?

```{r}
par(mfrow=c(2,2))
# setup plot ranges noting max of normal density is at mean
xrange <- c(1500,6000)
yrange <- range( c(n, area*dnorm(birth.mean, birth.mean, birth.sd), 0))
scale=0.6
par(mar=c(5,0,1,0)+0.1)

# (a) empirical distribution and fitted normal
plot(xrange, yrange, type = "n", xlab = "", ylab = "",
     bty="n",axes=F,main="(a) Distribution of birthweights", cex=scale)
axis(1,cex=scale) 
# draw bars using rect and density using curve
rect(weights, 0, weights - 500, n, col = "lightblue")
curve(area*dnorm(x, birth.mean, birth.sd), min(xrange), max(xrange), add = TRUE, 
      lwd=3, col="blue")
lines(c(xw,xw),yrange,col="red",lwd=2)
```

Below, I will chart the normal distribution with Standard Deviations.

Mathematically, around 95% of the area under the chart is between -2 and +2 SDs, and 99.8% between -3 and +3 SDs

Now, another way to think about this is that the area under the normal curve represents 100% of the probability of a given value (here, weight) occuring. So, you could say that it is around 95% probable that any given birthweight (randomly sampled) will occur between -2 and +2 SDs of the mean.

We can calculate this easily, given that the mean is 3345.263, and the SD is 577.7594

So, it is 95% probable that a random baby born in the UK in 2020 will weigh between 2189.745 and 4500.781g

```{r}
# (b)   plot with sds  
plot(xrange, yrange, type = "n", xlab = "", ylab = "",
     bty="n",axes=F,,main="(b) Mean +/- 1, 2, 3 SDs" )
axis(1)
curve(area*dnorm(x, birth.mean, birth.sd), min(xrange), max(xrange), add = TRUE, lwd=3, col="blue")
I=-3:3
x1=birth.mean+I*birth.sd
y1=area*dnorm(x1,birth.mean, birth.sd)
label=c("-3 SDs", "-2 SDs", "-1 SD", "mean", "+1 SD","+2 SDs", "+3 SDs")
bit=10000
xx=250
shift=c(-xx,-xx,-xx,0,xx,xx,xx)
for(i in 1:7){
  lines(c(x1[i],x1[i]), c(0,y1[i]),lwd=2)
  text(x1[i]+shift[i],y1[i]+bit,label[i],cex=0.75)
}
lines(c(xw,xw),yrange,col="red",lwd=2)
```

We can calculate how many SDs Monty's weight was above the mean by calculating the z-score for his weight of 3912g.

To do so, let's return to the slide deck and see the calculation.

***Move back to the slides...***

Next, we can overlay percentiles in 5% increments on the chart, which are also useful ways to describe a distribution like the Normal.

50th is of course the median, and the 25th and 75th percentilies are the quartiles.

Here, the same things we have previously discussed in relation to samples are applied to populations

We can also see that Monty lies on the 83rd percentile, that is 83% of babies born in 2020 weighed less. This tallies with what we were told at the time, and then things are monitored quite closely from that time on in terms of percentiles.

```{r}
plot(xrange, yrange, type = "n", xlab = "Birthweight (gms)", ylab = "",
     bty="n",axes=F,,main="(c) Percentiles" )
axis(1)
curve(area*dnorm(x, birth.mean, birth.sd), min(xrange), max(xrange), add = TRUE,
      lwd=3, col="blue")
I=c(1,5,25,50,75,95,99)
x1=qnorm(I/100, birth.mean,birth.sd)
y1=area*dnorm(x1,birth.mean, birth.sd)
label=c("1%", "5%", "25%", "50%","75%","95%","99%")
bit=5000
for(i in 1:7){
  lines(c(x1[i],x1[i]), c(0,y1[i]),lwd=2,lty=2)
  text(x1[i],-bit,label[i],cex=0.6)
}
lines(c(xw,xw),yrange,col="red",lwd=2)

```

Moving to a discussion of low birthweight. It is medically defined as low if below 2500g. Nuffield reports that 6.6% of babies in 2020 were born at low birthweight, which is fairly good.

This tallies fairly well with the predicted proportion from the normal distribution of 7.2%

What we gain from the grey shaded area is the proportion of the population which is expected to be born at low birthweight, but also an indication of the probability that a randomly-chosen baby born in 2020 would be of low weight.

This concept can easily be transferred to other weights in this population (and by extension to other variables and populations distributed like this)

```{r}
# (d)  Low birth weight  
plot(xrange, yrange, type = "n", xlab = "Birthweight (gms)", ylab = "",
     bty="n",axes=F,,main="(d) Low birth weight" )
axis(1)
curve(area*dnorm(x, birth.mean, birth.sd), min(xrange), max(xrange), add = TRUE,
      lwd=3, col="blue")
x1=seq(1500,2500,10)
y1=area*dnorm(x1,birth.mean, birth.sd)
polygon(c(x1,x1[101:1]),c(rep(0,101), y1[101:1]),col="lightblue")

lines(c(xw,xw),yrange,col="red",lwd=2)
x1=seq(1500,xw,10)
nx=length(x1)
y1=area*dnorm(x1,birth.mean, birth.sd)
polygon(c(x1,x1[nx:1]),c(rep(0,nx), y1[nx:1]),col="red",density=10)

text(2000,70000,"Proportion\n below\n 2500 gms\n = 7.2%",cex=0.75)
text(3400,70000,"Proportion\n below\n 3912 gms\n = 83.6%",cex=0.75)
```
