# T-Tests for Means {#t-test}

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
library(readxl)
library(ggplot2)
library(boot)
library(rstatix)
library(skimr)
library(infer)
library(ggpubr)
library(dplyr)
library(psych)
```

This Chapter supplements the material on t-tests in the second part of Lecture 6.

In this chapter, I will introduce the t-test, a classic analysis technique that is used to compare two groups together.

## Independent Sample T-Tests

We can also use bootstrapping as an entry point to a new analysis situation, where we are comparing two groups. This could be for example in a classic experimental context; treatment and control.

Remember, t-tests can be done in any analysis setting, and it does not require bootstrapping at all. It just so happens that they are nicely explainable at this point.

So, we are going to analyze a set of data from my infamous Ed Sheeran Study^1^ - which would certainly win an Ignobel Prize if I were ever to do it in reality rather than in my fondest imaginings.

The design was discussed in the lecture slides, but in brief we have a two-group 'treatment and control' experiment, the classic 'randomized control trial' style study.

The first thing to do is read in the data and describe it.

```{r}
ED_IND<-read_excel("Data/SHEERAN_T.xlsx", sheet = "IND")
```

```{r}
head(ED_IND)
describe(ED_IND)
ED_IND$GROUP <- factor(ED_IND$GROUP)
skim(ED_IND)
```

We can see that we have one FACTOR variable, which we need to indicate the group each subject was in (CONTROL being GROUP 1, and ED being GROUP 2).

So, let's run an *independent samples T-Test* with bootstrapped confidence interval.

We use an independent samples test, as the *theory* is these two groups - after participating in the experiment - are no longer the 'same'. There is something 'different' about the population of people who listen to Ed Sheeran, compared to those who do not. What we are trying to do is assess whether this difference is in *anger.*

Before we get to the code, it's worth pausing on what we're actually bootstrapping, and why it isn't simply the difference in the two group means.

The raw mean difference — a bit over 1 point on our five-point anger scale — is perfectly interpretable in its own right: it's a real number in the actual units the scale was measured in. But its size alone tells us nothing about how impressed we should be by it, because that depends entirely on the scale we happened to use and how much natural variation there already is in anger scores. A difference of "1 point" is enormous on a scale that barely moves and unremarkable on one that swings wildly.

This is the same problem as the standardised versus unstandardised regression estimates back in Section \@ref(assoc-rel). An unstandardised coefficient is interpretable in real-world units — cigarettes, miles cycled — but can't be compared across variables measured on different scales; standardising puts everything onto a common footing. The t-statistic does the same job here: it takes the mean difference and re-expresses it relative to how much variability and how much data we have, which is what lets us judge whether a given difference is impressively large or just noise.

```{r}

#First, some code to display the group means using dplyr
ED_IND %>%   # Specify data frame
group_by(GROUP) %>% # Specify group indicator
summarise_at(vars(ANGER), # Specify column
list(name = mean))    # Specify function

#use infer to bootstrap the t-value
#https://infer.netlify.app/articles/observed_stat_examples

t_hat <- ED_IND %>%
  specify(ANGER ~ GROUP) %>%
  calculate(stat = "t", order = c("1", "2"))

boot_dist <- ED_IND %>%
   specify(ANGER ~ GROUP) %>%
   generate(reps = 1000, type = "bootstrap") %>%
   calculate(stat = "t", order = c("1", "2"))

percentile_ci <- get_ci(boot_dist)


visualize(boot_dist) +
  shade_confidence_interval(endpoints = percentile_ci)

#display the t and CI
t_hat
percentile_ci

```

Remember: Group 1 is the control, and Group 2 listened to Ed Sheeran.

Cool, so it seems that Group 2 displayed more anger. The confidence interval for the t-statistic does not contain 0, so it supports the idea that there is a difference here. The interval is quite wide though - because of our small sample size. In a later Chapter we'll return to this issue of what confidence intervals actually mean.

I'll be honest: in a purely bootstrap context like this one, that standardisation step can feel like an unnecessary detour. We didn't need the t to reach our conclusion — the conclusion was always going to be "does the confidence interval contain zero or not", and that would have been exactly as true if we'd bootstrapped the raw mean difference directly instead. The t only really earns its keep once we start relying on known probability distributions rather than brute-force resampling, which is where we're headed in the next chapter. For now, take it as scaffolding for what's coming rather than something bootstrapping strictly requires.

## Paired-Sample T-tests

OK, so let's use a different design, using a paired samples t-test. Let me go *back to the slides...* to explain this difference.

The same basic process is needed, but with some modifications because of the type of comparison we are doing. And, as such, we have some new data.

```{r}
ED_PAIR<-read_excel("Data/SHEERAN_T.xlsx", sheet = "PAIR")
ED_PAIR <- data.frame(ED_PAIR)
head(ED_PAIR)
describe(ED_PAIR)
```

Now, for this bootstrap purpose we actually need to to calculate the **difference between the two measurements** (here, T1 and T2).

Then, we bootstrap a one-sample t-test with this difference variable. This essentially tests whether the difference is different from zero. If it is, that indicates some effect of the treatment.

Let us first calculate the new variable:

```{r}
ED_PAIR$DIF <- ED_PAIR$ANG_T2-ED_PAIR$ANG_T1
head(ED_PAIR)
describe(ED_PAIR)
```

We can see there is a new 'DIF' variable here.

Notice we're bootstrapping the mean of this difference directly, not a t-statistic as we did for the independent-samples design above. It's the same underlying question — is this far from zero? — just reached without the standardisation step in between. Nothing is lost by skipping it: with paired data there's a single already-differenced variable rather than two separately-scaled groups to compare, so there's less need to standardise before asking the question.

Next, we bootstrap a confidence interval for the *mean of the difference variable*, to see whether it includes zero:

```{r}
#modified from https://infer.netlify.app/articles/observed_stat_examples.html


x_bar <- ED_PAIR %>% 
  specify(response = DIF) %>%
  calculate(stat = "mean")

boot_dist <- ED_PAIR %>%
   specify(response = DIF) %>%
     generate(reps = 1000, type = "bootstrap") %>%
   calculate(stat = "mean")

percentile_ci <- get_ci(boot_dist)

visualize(boot_dist) +
  shade_confidence_interval(endpoints = percentile_ci)

#report the results
x_bar
percentile_ci
mean(ED_PAIR$ANG_T1)
mean(ED_PAIR$ANG_T2)


```

We can do this in another way, but instead running a 1-sample t-test on the difference variable, testing whether it is different from zero:

```{r}
#modified from https://cran.r-project.org/web/packages/infer/vignettes/observed_stat_examples.html
x_bar <- ED_PAIR %>% 
  specify(response = DIF) %>%
  calculate(stat = "mean")

null_dist <- ED_PAIR %>%
   specify(response = DIF) %>%
  hypothesize(null="point", mu=0)%>% #this is the additional bit
   generate(reps = 1000, type = "bootstrap") %>%
   calculate(stat = "mean")

percentile_ci <- get_ci(boot_dist)

visualize(null_dist) +
    shade_p_value(obs_stat = x_bar, direction = "two-sided")

null_dist %>%
  get_p_value(obs_stat = x_bar, direction = "two-sided")

x_bar

mean(ED_PAIR$ANG_T1)
mean(ED_PAIR$ANG_T2)


```

Marvelous. We can see that the results from both variants of this test suggest that after listening to Ed Sheeran, our sample on average reported more anger. 

1: Because the 95% confidence interval does not include 0, I am confident in saying that there is some effect going on here.

2: Because the bootstrapped probability (as expressed by the p-value) of observing the mean difference in anger between T1 and T2 of 1.33 is very low (0 in our results, but the true value will not be exactly zero), I am confident in saying there is an effect here.

Not surprised...

^1^If you are a fan, friend, or relative of, or more importantly a lawyer for, Ed Sheeran, please note that this is not a real study. Furthermore, I am sure Ed Sheeran is a very nice person and I would delight in his company. This is merely a bit of humour. 
