# T-Tests for Means {#t-test}

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
library(readxl)
library(ggplot2)
library(boot)
library(rstatix)
library(skimr)
library(infer)
library(ggpubr)
library(dplyr)
library(psych)
```

This Chapter supplements the material on t-tests in the second part of Lecture 6.

In this chapter, I will introduce the t-test, a classic analysis technique that is used to compare two groups together.

## The Christmas Music Study

We can also use bootstrapping as an entry point to a new analysis situation, where we are comparing two groups. This could be for example in a classic experimental context; treatment and control.

Remember, t-tests can be done in any analysis setting, and it does not require bootstrapping at all. It just so happens that they are nicely explainable at this point.

So, here's the study I'll use for the rest of this chapter, and pick up again a couple of chapters from now. Participants come in and proofread a passage of text — a genuinely tedious task, by design — with background audio playing. I tell them the study is about concentration and background noise, which is true as far as it goes, but isn't the whole story. Half the participants get silence, half get Christmas music (which is, delightfully, one of the few things guaranteed to mildly irritate almost everybody, making it a gift for a researcher even if not for the participants). Before and after the task, I measure how irritated people feel, using four items — irritated / annoyed / tense / on edge, each rated 1-7 — and average them into a single composite score, `IRRIT`. The four items hold together well as one measure: Cronbach's alpha across them is a healthy 0.86 in this sample, so treating them as a single composite rather than four separate numbers is a reasonable thing to do.

Before looking at any numbers, there's a design question worth sitting with, because it's the reason this chapter runs two different-looking t-tests on what can feel like the same question.

Suppose the Christmas-music group's irritation goes up from before the task to after it. Is that enough, on its own, to conclude Christmas music is irritating? No — because the task itself is boring. Twenty minutes of proofreading a dull passage would probably wind most people up a bit regardless of what's playing in the background, including nothing at all. A rise from baseline to post-task doesn't tell you the *music* did anything, because the task alone might have. What actually answers that question is a comparison group who did the identical boring task in silence, so I can see whether the Christmas-music group ended up *more* irritated than boredom alone would predict. That's why this chapter needs both an independent-samples comparison (Christmas vs silence, after the task) and a paired comparison (the Christmas group's own before-and-after) — they answer two different questions, and only one of them can rule out "it was just the task."

## Independent-Samples T-Test

The first thing to do is read in the data and describe it.

```{r}
ED_IND<-read_excel("Data/MUSIC_STUDY.xlsx", sheet = "S1_IND")
```

```{r}
head(ED_IND)
describe(ED_IND)
ED_IND$GROUP <- factor(ED_IND$GROUP, levels = c("Silence", "Christmas"))
skim(ED_IND)
```

We can see that we have one FACTOR variable, which indicates which group each subject was in — Silence or Christmas.

So, let's run an *independent samples T-Test* with bootstrapped confidence interval, comparing the two groups' irritation scores at the end of the task (T2).

We use an independent samples test here because these are two different groups of people, who did the task under two different conditions. What we're trying to assess is whether the Christmas-music group ended up more irritated than the silence group.

Before we get to the code, it's worth pausing on what we're actually bootstrapping, and why it isn't simply the difference in the two group means.

The raw mean difference — a bit over 1.5 points on our seven-point irritation scale — is perfectly interpretable in its own right: it's a real number in the actual units the scale was measured in. But its size alone tells us nothing about how impressed we should be by it, because that depends entirely on the scale we happened to use and how much natural variation there already is in irritation scores. A difference of "1.5 points" is enormous on a scale that barely moves and unremarkable on one that swings wildly.

This is the same problem as the standardised versus unstandardised regression estimates back in Section \@ref(assoc-rel). An unstandardised coefficient is interpretable in real-world units — cigarettes, miles cycled — but can't be compared across variables measured on different scales; standardising puts everything onto a common footing. The t-statistic does something similar here: it takes the mean difference and re-expresses it relative to how much variability and how much data we have, which is what lets us judge whether a given difference stands out from the noise. It isn't quite the same thing as standardising, though, and I'll come back to why at the end of this section.

```{r}

#First, some code to display the group means using dplyr
ED_IND %>%   # Specify data frame
group_by(GROUP) %>% # Specify group indicator
summarise_at(vars(IRRIT), # Specify column
list(name = mean))    # Specify function

#use infer to bootstrap the t-value
#https://infer.netlify.app/articles/observed_stat_examples.html

t_hat <- ED_IND %>%
  specify(IRRIT ~ GROUP) %>%
  calculate(stat = "t", order = c("Christmas", "Silence"))

boot_dist <- ED_IND %>%
   specify(IRRIT ~ GROUP) %>%
   generate(reps = 1000, type = "bootstrap") %>%
   calculate(stat = "t", order = c("Christmas", "Silence"))

percentile_ci <- get_ci(boot_dist)


visualize(boot_dist) +
  shade_confidence_interval(endpoints = percentile_ci)

#display the t and CI
t_hat
percentile_ci

```

Cool, so it seems that the Christmas-music group displayed more irritation. The confidence interval for the t-statistic does not contain 0, so it supports the idea that there is a real difference here. The interval is quite wide though - because of our small sample size. In a later Chapter we'll return to this issue of what confidence intervals actually mean.

I'll be honest: in a purely bootstrap context like this one, that standardisation step can feel like an unnecessary detour. We didn't need the t to reach our conclusion — the conclusion was always going to be "does the confidence interval contain zero or not", and that would have been exactly as true if we'd bootstrapped the raw mean difference directly instead. The t only really earns its keep once we start relying on known probability distributions rather than brute-force resampling, which is where we're headed in the next chapter. For now, take it as scaffolding for what's coming rather than something bootstrapping strictly requires.

```{r t-test-cohens-d, include=FALSE}
irrit_means <- tapply(ED_IND$IRRIT, ED_IND$GROUP, mean)
irrit_sds   <- tapply(ED_IND$IRRIT, ED_IND$GROUP, sd)
irrit_ns    <- table(ED_IND$GROUP)
irrit_sp    <- sqrt(((irrit_ns[["Silence"]] - 1) * irrit_sds[["Silence"]]^2 +
                     (irrit_ns[["Christmas"]] - 1) * irrit_sds[["Christmas"]]^2) /
                    (sum(irrit_ns) - 2))
irrit_d     <- (irrit_means[["Christmas"]] - irrit_means[["Silence"]]) / irrit_sp
irrit_d_label <- ifelse(abs(irrit_d) >= 0.8, "large",
                   ifelse(abs(irrit_d) >= 0.5, "medium",
                     ifelse(abs(irrit_d) >= 0.2, "small", "smaller than small")))
irrit_n_per_group <- min(irrit_ns)
```

There's one more way of putting that difference on a common footing, and it's the one you'll meet most often in published papers. The t-statistic divides the mean difference by its standard error, and the standard error shrinks as the sample grows. That makes t good at answering "does this difference stand out from the noise?" but not "how big is it?" Run exactly the same study with ten times as many people, get exactly the same difference, and the t would be roughly three times larger.

*Cohen's d* answers the second question. It divides the mean difference by the standard deviation of the irritation scores themselves, pooled across the two groups, so it tells you how many standard deviations apart the groups ended up. Unlike t, it doesn't grow just because the sample does. For our study, d is `r round(irrit_d, 2)`: the Christmas-music group finished about `r round(irrit_d, 1)` standard deviations more irritated than the silence group. It's a direct cousin of the standardised regression coefficients in Chapter \@ref(assoc-rel): the same idea in a different setting.

You'll often see d described using Jacob Cohen's labels, where 0.2 is "small," 0.5 "medium" and 0.8 "large." By those labels ours is `r irrit_d_label`. But Cohen offered them as a convention for use when nothing better suggests itself [@cohen1988statistical], and they've since hardened into exactly the kind of rule this book keeps warning about. Two cautions apply to our own number, too. First, with only `r irrit_n_per_group` people per group, it's a noisy estimate, and as you'll see in Chapter \@ref(Issues), big-looking effects from small samples deserve particular scepticism. Second, like any standardised effect, d tells you nothing about whether the difference matters in practice. For that you have to go back to the original units, which here means a difference of `r round(irrit_means[["Christmas"]] - irrit_means[["Silence"]], 1)` points on a seven-point scale.

## Paired-Sample T-test

OK, so let's use a different design on the same study — the Christmas-music group's own scores, before the task (T1) and after it (T2) — using a paired samples t-test. Let me go *back to the slides...* to explain this difference.

The same basic process is needed, but with some modifications because of the type of comparison we are doing. And, as such, we have some new data.

```{r}
ED_PAIR<-read_excel("Data/MUSIC_STUDY.xlsx", sheet = "S1_PAIR")
ED_PAIR <- data.frame(ED_PAIR)
head(ED_PAIR)
describe(ED_PAIR)
```

Now, for this bootstrap purpose we actually need to to calculate the **difference between the two measurements** (here, T1 and T2).

Then, we bootstrap a one-sample t-test with this difference variable. This essentially tests whether the difference is different from zero. If it is, that indicates some effect of the treatment.

Let us first calculate the new variable:

```{r}
ED_PAIR$DIF <- ED_PAIR$IRRIT_T2-ED_PAIR$IRRIT_T1
head(ED_PAIR)
describe(ED_PAIR)
```

We can see there is a new 'DIF' variable here.

Notice we're bootstrapping the mean of this difference directly, not a t-statistic as we did for the independent-samples design above. It's the same underlying question — is this far from zero? — just reached without the standardisation step in between. Nothing is lost by skipping it: with paired data there's a single already-differenced variable rather than two separately-scaled groups to compare, so there's less need to standardise before asking the question.

Next, we bootstrap a confidence interval for the *mean of the difference variable*, to see whether it includes zero:

```{r}
#modified from https://infer.netlify.app/articles/observed_stat_examples.html


x_bar <- ED_PAIR %>%
  specify(response = DIF) %>%
  calculate(stat = "mean")

boot_dist <- ED_PAIR %>%
   specify(response = DIF) %>%
     generate(reps = 1000, type = "bootstrap") %>%
   calculate(stat = "mean")

percentile_ci <- get_ci(boot_dist)

visualize(boot_dist) +
  shade_confidence_interval(endpoints = percentile_ci)

#report the results
x_bar
percentile_ci
mean(ED_PAIR$IRRIT_T1)
mean(ED_PAIR$IRRIT_T2)


```

We can do this in another way, but instead running a 1-sample t-test on the difference variable, testing whether it is different from zero:

```{r}
#modified from https://cran.r-project.org/web/packages/infer/vignettes/observed_stat_examples.html
x_bar <- ED_PAIR %>%
  specify(response = DIF) %>%
  calculate(stat = "mean")

null_dist <- ED_PAIR %>%
   specify(response = DIF) %>%
  hypothesize(null="point", mu=0)%>% #this is the additional bit
   generate(reps = 1000, type = "bootstrap") %>%
   calculate(stat = "mean")

percentile_ci <- get_ci(boot_dist)

visualize(null_dist) +
    shade_p_value(obs_stat = x_bar, direction = "two-sided")

null_dist %>%
  get_p_value(obs_stat = x_bar, direction = "two-sided")

x_bar

mean(ED_PAIR$IRRIT_T1)
mean(ED_PAIR$IRRIT_T2)


```

Marvelous. We can see that the results from both variants of this test suggest that after the proofreading task with Christmas music playing, our sample on average reported more irritation than they did beforehand.

1: Because the 95% confidence interval does not include 0, I am confident in saying that there is some effect going on here.

2: Because the bootstrapped probability (as expressed by the p-value) of observing the mean difference in irritation between T1 and T2 of 1.7 is very low (0 in our results, but the true value will not be exactly zero), I am confident in saying there is an effect here.

Not surprised — but remember the design question from the start of the chapter: this rise, on its own, doesn't tell us the *music* did it, because the boring task alone could have produced some of it. That's exactly what the independent-samples comparison above was for, and it agrees — the Christmas-music group ended up more irritated than the silence group who did the identical task.

But there's a limit to what even that comparison can tell us. All I've actually shown is that background music of *some* kind moves irritation more than no music at all — I haven't shown there's anything special about *Christmas* music specifically, as opposed to background music in general. The design as it stands can't separate "background music is irritating" from "Christmas music in particular is irritating," because silence is the only comparison group I have. That's a real limitation, not a minor caveat, and it's exactly the question Chapter \@ref(ANOVA) picks up, with a third group added to settle it. It's also exactly the confounding problem from Chapter \@ref(causation): background music and Christmas music are tangled together in this design, so I can't yet credit one over the other.

## Reading the literature {#t-test-reading}

Comparing two groups is the simplest design there is, and it turns up in management papers far more often than the phrase "t-test" does. Sometimes you'll see a t-test reported as such, usually in an experiment comparing a treatment with a control. Much more often, the comparison is hiding inside a regression as a single coefficient. Either way, the questions from this chapter apply: what exactly is being compared with what, how big is the difference, and could something other than the treatment have produced it?

### Where you'll meet this

**Reported t-tests.** In experimental papers you'll see something like t(58) = 2.41, p = .019. The number in brackets is the degrees of freedom, which for two independent groups is the total sample size minus two. If it's a decimal, such as t(41.7), the authors have used *Welch's t-test*, which doesn't assume the two groups are equally variable. That's a good sign rather than an oddity: there's a strong case for using Welch's version by default [@delacre2017why].

**Group differences as regression coefficients.** A classic t-test comparing two groups gives exactly the same answer as a regression with a single *dummy variable*, coded 1 for one group and 0 for the other. The coefficient on the dummy is the difference between the two group means, and its t-value is the t-test. So when a paper reports a coefficient for "female," "family firm" or "treated," it's telling you the difference between two group means, adjusted for whatever else is in the model. Two things follow. First, you need to know which group was coded 0, because the coefficient is always relative to that *reference category*: flip the coding and the sign flips. With more than two categories, such as industries or regions, every coefficient compares one category with the one left out, and the choice of which to leave out decides which comparisons you get to see. Second, "adjusted for" is doing a lot of work, for all the reasons in Chapter \@ref(causation).

**Cohen's d and its relatives.** You'll also meet *Hedges' g*, which applies a small correction that matters in small samples. It's fine to read the two as essentially the same thing. The trap is in before-and-after designs like the paired test in this chapter. There, d can be calculated by dividing by the standard deviation of the *differences* rather than of the scores. When people's before and after scores are strongly related, as they usually are, that standard deviation is small, and the resulting d can be much larger than an independent-groups d for the same difference [@lakens2013calculating]. So a d from a within-person study isn't necessarily comparable with one from a between-groups study, and a paper comparing its effect with "typical" effects elsewhere should say which kind it calculated. And like every standardised effect, d needs translating back into natural units before you know whether it matters, as Section \@ref(intro-reading) argued.

### The connection back

Every group comparison in a paper, whether it's labelled a t-test or hidden in a coefficient, comes down to the two questions this chapter asked about the Christmas music study. Is the difference bigger than chance alone would produce? And is the comparison group the right one, so that the difference can be put down to the thing the authors say caused it? Statistics can answer the first question. Only the design can answer the second.

### Questions worth asking

1. **What exactly is being compared with what?** Which group is the reference category?
2. **What would the comparison group need to be to rule out the most obvious alternative explanation?** Is it?
3. **How big is the difference in the original units, as well as in standard deviations?**
4. **If it's a within-person design, how was the effect size calculated?** Is it being compared with effects calculated the same way?

### One red flag

Beware the before-and-after study with no comparison group. A training programme, an incentive scheme or a culture initiative is introduced, scores are measured before and after, and they go up. The paired test is significant, and the paper concludes that the intervention worked. But as the Christmas music study showed, a rise from before to after can't tell you on its own what caused it. People change over time anyway, organisations change around them, and those who start with unusually low scores tend to score closer to average next time regardless. Without a group that went through the same period without the intervention, the rise describes what happened. It doesn't explain it.

### If you want to read more

Daniël Lakens's primer on calculating and reporting effect sizes [@lakens2013calculating] is free to read online, practical, and especially good on the different versions of d and when each one is appropriate. It's written for researchers rather than readers, but it's the clearest account I know of what the numbers in an effect-size table actually mean.

<!-- APPENDIX POINTER: uncomment once the reader-guide appendix exists.
The questions in this section also appear, in the order you'd meet them when reading a paper, in Appendix \@ref(reader-guide).
-->
