# Classical Statistical Hypothesis Testing {#H-testing}

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(readxl)
library(ggplot2)
library(ggpubr)
library(mosaic)
library(dplyr)
library(skimr)
library(rstatix)
library(pwr)
```

This Chapter relates to the first part of Lecture 8, and contains some (re)worked examples from earlier chapters / lessons, focusing specifically on the hypothesis testing aspect of them.

## Testing a Real Hypothesis: Are WBS PhD Salaries Different? {#salary-ztest}

Back in Section \@ref(conf) we used the WBS PhD graduate salary data — 42 graduates, mean starting salary $170,000, standard deviation $46,617.4 — to build a confidence interval. That was useful, but it didn't actually answer the question I cared about when I started looking at this data: can I claim, in the PhD program prospectus, that WBS graduates earn more than graduates elsewhere?

A confidence interval alone doesn't let me say that. It tells me something about precision, not about a comparison. To make a comparison I need a number to compare against, and I need to state, in advance, how different my sample would have to be before I'm willing to call it a real difference rather than the ordinary noise you'd expect from any sample of 42 people.

That comparison number is my null hypothesis, and I didn't invent it out of thin air. The American Marketing Association runs a doctoral special interest group, and every year they survey new marketing PhD graduates about starting salary and send me the results — mostly, I suspect, to depress me about what I earned in my first job. The most recent figure was US$145,000. It's not a perfect number (a proper job would involve going round every PhD director in the business school and averaging across fields), but it's a real, defensible estimate of "what graduates elsewhere earn", which is exactly what a null hypothesis needs to be.

- **Null hypothesis:** the average starting salary of WBS PhD graduates is not different from US$145,000.
- **Sample:** 42 WBS graduates, sample mean $170,000, sample SD $46,617.4 — the same sample as Section \@ref(conf).

Before doing any calculation, there's a decision to make, and it's the decision that usually gets made by accident, halfway through the arithmetic, once everyone's attention is already used up. Did I hypothesize a direction? I could have framed the alternative as directional — WBS graduates earn *more* than $145,000 — or non-directional — WBS graduates earn a *different* amount, which could be more or less. A directional hypothesis is the stronger claim, and it has to be committed to in advance, not read off the result after the fact.

Here's the honest version: privately, of course I was expecting WBS graduates to earn more. But my null hypothesis, as written, was "not different from $145,000", which commits me to the non-directional alternative whatever I was quietly hoping. I didn't encode the direction I actually believed in, so I don't get to use it. That's not me being pedantic about my own case for the sake of the example — it's the actual mechanism. A lot of people hypothesize a direction in their head and then run a two-tailed test anyway, mostly because no one ever made them sit down and decide properly. Here, the test is two-tailed.

With that settled, the arithmetic is almost an anticlimax. What I want to know is: if the population mean really is $145,000, how far away — in standardised units — is my observed sample mean of $170,000?

```{r}
null.mean <- 145000
sample.mean <- 170000
sample.sd <- 46617.4
n <- 42

se <- sample.sd / sqrt(n)
z <- (sample.mean - null.mean) / se
z
```

Same standardisation logic as everywhere else in this book: take what I observed, subtract what I'd expect under the null, divide by how much variation sampling alone would produce. The effect is to redraw the whole problem on a scale where the null sits at zero, and my sample mean lands wherever it lands. A z of around 3.5 means my sample mean is roughly three and a half standard errors out from where the null says it should be — well beyond the 1.96 we've used as our 95% cutoff elsewhere.

```{r}
p.two.tailed <- 2 * (1 - pnorm(abs(z)))
p.two.tailed
```

That p-value — somewhere around 0.0005, about 1 in 2,000 — means: *if* the null is true, the probability of drawing a sample of 42 with a mean this far out, in either direction, is about 1 in 2,000. It does not mean there's a 1-in-2,000 chance the null is true. A p-value says nothing about the probability of any hypothesis. It's a statement about how surprising the data would be under an assumption, not a verdict on the assumption.

So, do I reject the null? This is a judgement, not a law of nature. Nothing about this p-value compels a conclusion — I could, in principle, decide this is just an unusually extreme sample from a population that really does average $145,000, and nothing in the maths would stop me. What I actually do, like almost everyone else, is apply a cutoff — conventionally p < 0.05 — and treat anything below it as evidence against the null. That cutoff exists because people like rules, not because 0.05 has any special status in nature. Here the data are miles past any conventional threshold, so the call is easy. It won't always be, and when it isn't, the arithmetic still won't make the decision for you.

**Directional vs non-directional, side by side:**

| | One-tailed (directional) | Two-tailed (non-directional) |
|---|---|---|
| Alternative hypothesis | States a direction — e.g. "bigger than $145,000" | States only "different from $145,000" |
| Must be decided | Before looking at the data | Before looking at the data |
| Relationship between the p-values | Half the two-tailed p, for the same z | Double the one-tailed p, for the same z |
| This example | Not what I actually committed to | What I actually committed to |

## Premier League Goals and the Poisson Distribution

In this section, we are not really worried about the *actual* goals or anything like that. What we are concerned about is formally testing the hypothesis that Premier League goals per game actually do follow a Poisson distribution. To do this, we will use what is called a **chi-square** test, which uses the chi-square statistic.

Chi-square is a fantastically versatile statistic, which is often used to compare tables of observations. Commonly, we compare a table of observed events with a table of events that we would expect if the null hypothesis were true in the population. The chi-square value is the measure of the difference between observed and expected counts in the table. We can use it here to compare the amount of goals we actually did observe in the 21-22 season, with those that we would *expect* to observe if the goals per game really do follow a Poisson distribution.

Again, much of the idea and code for this section comes from: <https://bookdown.org/theqdata/honors_thesis/goal-scoring-and-the-poisson-process.html>

The data comes from: <https://www.football-data.co.uk/englandm.php> and is the EPL results for 21-22 Season, with a new column of Total Goals (TOTG) created by me in Excel.

```{r}
EPL<-read_excel("Data/EPL21-22.xlsx")
head(EPL)
```

```{r}
summary(EPL$HomeTeam)
```

The above is a quick way of calculating / checking the total number of games if you don't know already.

```{r}
summary(EPL$TOTG)
```

```{r}
sum(EPL$TOTG)
```

First, let's create a table of all the matches with different numbers of goals, just as we did in Chapter \@ref(probability).

```{r}
GoalsTable <- 
  EPL %>% 
  group_by(TOTG) %>% 
  summarise(ActualMatches = n())
GoalsTable 
```

Here, we have a problem. The chi-square test has a condition that the expected values in every column will be greater than 5, and as we go down the Poisson probabilities (i.e. for higher total goals), it is clear that this will start to become a problem for us.

The solution is simple, we combine the numbers for larger numbers of goals into a single 'x and above' variable. Here, let's do it for 6 goals and above.

To do so, we first need to create a new table, where the values for 6 and above are combined into a new row.

```{r}
# select first 5 rows (0, 1, 2, 3, 4, 5 goals)
NewGoalsTable <- GoalsTable[1:6,]
# sum up the remaining rows
NewGoalsTable[7,] <- sum(GoalsTable[7:nrow(GoalsTable),2])
NewGoalsTable <- mutate(NewGoalsTable, TOTG = as.character(TOTG))
# put in 1 category called "5 or more"
NewGoalsTable[7,"TOTG"] <- "6 or more" 
NewGoalsTable
```

Now we can create the Poisson distribution for the relevant mean, which means we first need to create the stats...

```{r}
fav_stats(EPL$TOTG)
```

Below is some code which will help us to create the distributions and also check them.

```{r}
MeanGoals <- fav_stats(EPL$TOTG)[[6]]
numMatches <- fav_stats(EPL$TOTG)[[8]]
StDevGoals <- fav_stats(EPL$TOTG)[[7]]
VarianceGoals <- StDevGoals ^ 2
```

```{r}
MeanGoals
```

```{r}
VarianceGoals
```

As we saw in Chapter \@ref(probability), while for a Poisson distribution these should be exactly the same, we decided that they were good enough to be going on with.

So, let's actually check this formally with the Chi-square test.

We do so by first creating a table of the Poisson probabilities for x or fewer goals, for the Poisson with a mean of 2.818 (i.e. the variable MeanGoals)

```{r}
PoisProb <- dpois(c(0:6), MeanGoals)
POIS <-data.frame(PoisProb)
POIS
```

This allows us to make some predictions of what we could expect.

Remembering again that there are 380 games in a season, we can see that there is a 0.16 chance of seeing 4 goals in a game, which equates to 380 x 0.16 = 60.8 games we would expect to see in a season with 4 goals.

We can check this out by creating a new table comparing actual with predicted values...

```{r}
NewGoalsTable <- cbind(NewGoalsTable, PoisProb) 
NewGoalsTable <- mutate(NewGoalsTable, 
                        ExpectedMatches = round(numMatches * PoisProb))
NewGoalsTable
```

This looks quite frighteningly close up until you get to the 5 row. It will be interesting to see how we go with a formal chi-square test.

```{r}
TOTGChisq <- chisq.test(NewGoalsTable$ActualMatches, 
                      p = NewGoalsTable$PoisProb, rescale.p = TRUE)
TOTGChisq
```

The P-value here is actually 0.64. This suggests that there is not a significant difference between the distribution of the actual goals, and that which would be expected if they did follow a Poisson distribution.

## Correlation Significance Tests

Here, we revisit our correlation between GDP per capita, and Happiness metrics, which I pulled from Our Word in Data:

```{r}
Happy<-read_excel("Data/HappyGDP.xlsx", sheet = "2020")

summary(Happy)
head(Happy)
```

Let's not worry about plotting the data, and go straight to the correlation:

```{r}
Assoc1 <- cor.test(Happy$Happiness, Happy$GDPpc, 
                    method = "pearson")

Assoc1

```

Here are our results. The estimate is a correlation, and we test that using the **t statistic**. The *t-value* is simply the estimate divided by the standard error (which we can't see in this output), and is interpreted essentially as 'how far from 0 is the estimate, in standard errors'.

The *p*-value for t is very very small, and obviously less than 0.05.

Conclusion - reject null hypothesis, accept alternative hypothesis (as always, pending better evidence).

Importantly, this does not mean that the **true correlation in the population** is 0.745, simply that it is *very unlikely to be zero*.

We can then look at our estimate of 0.745, and - even better - our confidence interval (see Section \@ref(conf) for information on how to interpret confidence intervals), to gain some indication of the likely true correlation in the population.

## Regression Significance Tests

The process to asses the significance of regression estimates is very very similar to that for correlations. Let's revisit the heart disease data set we used earlier.

```{r}
Heart<-read_excel("Data/heart.data.xlsx")

summary(Heart)
head(Heart)
```

Let's go straight to the multiple regression model.

```{r}
heart.disease.lm<-lm(heart.disease ~ biking + smoking, data = Heart)

summary(heart.disease.lm)
```

We interpret these just as we did the correlation significance tests.

The t-value is large, and the *p*-value (two-tailed) is small.

Interestingly, here we are given 'stars' for the different levels of significance, so to some extent the software is doing some decision making for you. To be honest, I always caution against relying solely on looking for 'stars' (it's actually a bit of a running joke amongst the people who have known me a long time that I once told an entire class in the 1990s to 'just look for the stars'). That's because the actual significance or not decision is based on the critical value and one- or two-tailed decision. The software often makes an assumption of 0.05 critical value for *p*, two-tailed, and calculates the 'stars' based on that. Sometimes that can conflict with the decision you have made yourself about what should be significant or not. That can trip you up if you didn't know to change these values in the software package.

Further, it also sort of entrenches the idea that things can be 'more' or 'less' statistically significant. Take a look at the results above, you'll see 3 stars represents a significance of '0', and 2 stars represents '0.001', boring old '0.05' only gets a single star, and '0.1' gets a dot. I'm not a fan here because this encourages the analyst to post-hoc make decisions about 'marginally' significant, or 'very' significant. These concepts do not actually exist in real statistical theory. Instead, you decide your critical value ***a priori***, and you either pass or fail it. As Yoda may have said if he was a statistician, "Pass or do not pass. There is no marginally significant".

Am I perfect? No. Specifically, do any of my papers use the language of 'marginal' significance? Sure, I bet you could find them. I am never a fan though, and I can promise you I argued about it at the time!
