# ANOVA {#ANOVA}

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(readxl)
library(ggplot2)
library(ggpubr)
library(mosaic)
library(dplyr)
library(rstatix)
library(pwr)
```

This Chapter provides analysis and examples related to the second part of Lecture 8.

Let's pick up exactly where Chapter \@ref(t-test) left off. That chapter showed that people who did a boring proofreading task with Christmas music playing ended up more irritated than people who did the identical task in silence — but it couldn't tell us whether that was really about *Christmas* music, or just about having background music of any kind playing while you're bored. Silence was the only comparison available, so the design couldn't separate the two explanations.

Study 2 fixes that by adding a third group: people who did the same task with background music that wasn't Christmas music, but was matched to it on tempo and familiarity, so it isn't a softer comparison to beat just because it's blander. Now there are three groups — Silence, Other (music), and Christmas — which finally lets us ask the question Chapter \@ref(t-test) left hanging: is it Christmas music specifically that's the problem, or just having something playing in the background at all? Randomly assigning people to one of the three groups is exactly the confound-neutralising move from Chapter \@ref(causation) — it's what lets this design pull the two explanations apart at all.

With three groups instead of two, running three separate t-tests would walk straight into the multiple comparisons problem (more on that in the next chapter). ANOVA instead tests whether there's a difference among all three groups with a single significance test, which is where we'll start.

```{r}
ED<-read_excel("Data/MUSIC_STUDY.xlsx", sheet = "S2_ANOVA")
ED$GROUP <- factor(ED$GROUP, levels = c("Silence", "Other", "Christmas"))
summary(ED)
head(ED)
```

Let's create a quick table of the group means, and visualise it with a simple boxplot of the groups.

```{r}
ED %>%
  group_by(GROUP) %>%
  get_summary_stats(IRRIT, type = "mean_sd")
ggboxplot(ED, x = "GROUP", y = "IRRIT")
```

Interesting picture. Silence is our baseline, Other is the matched background music without any Christmas content, and Christmas is, well, Christmas music. This is what we might call 'model-free' analysis. We are just looking at the data to see if intuitively it looks like we expect it to.

Let's run an ANOVA now, which places a statistical model on it, and tests whether or not the data supports the model.

```{r}
res.aov <- ED %>% anova_test(IRRIT ~ GROUP)
res.aov
```

Results here suggest there is a significant effect (p-value is very small). We also have an *effect size* measure ('ges', or generalised eta-squared) which is very useful to us, and suggests the effect is quite large. This can be interpreted in a similar way to R² in a regression: it's the proportion of the variance in the dependent variable (irritation) that is explained by group membership.

However, ANOVA only tests the *general effect* of the treatment / group. We don't know whether this is because of the difference between **all of the groups**, or only **some**. If it's background music *in general* that's irritating, we'd expect Silence to differ from both Other and Christmas, but Other and Christmas to look much the same as each other. If it's *Christmas music specifically*, we'd expect the opposite pattern: Christmas differs from both Silence and Other, but Silence and Other look much the same as each other. Only a post-hoc comparison of every pair can tell us which of these it actually is.

## Post-Hoc Testing

We can investigate this using **post-hoc tests**, which compare the difference between each pair of groups. There are many different types of post-hoc test for ANOVA, but the most typical one to use is the Tukey's test, which is what we will do here.

```{r}
pwc <- ED %>% tukey_hsd(IRRIT ~ GROUP)
pwc
```

We can actually plot these results in a really effective way:

```{r}
# Visualization: box plots with p-values
pwc <- pwc %>% add_xy_position(x = "GROUP")
ggboxplot(ED, x = "GROUP", y = "IRRIT") +
  stat_pvalue_manual(pwc, hide.ns = TRUE) +
  labs(
    subtitle = get_test_label(res.aov, detailed = TRUE),
    caption = get_pwc_label(pwc)
    )
```

This is the *Christmas-specifically* pattern from a moment ago: Christmas differs clearly from both Silence and Other, and Silence and Other don't differ significantly from each other (p = .118). Taken at face value, that looks like a clean answer — it really is something about Christmas music, not background music in general.

But look at the Silence-vs-Other comparison a bit more carefully, because this is exactly the case that a bare significant/not-significant reading undersells. Its 95% confidence interval is -0.12 to 1.28. That interval *only just* crosses zero — the lower bound is a hair's breadth into negative territory. Reading "not significant" as "proven: no difference" throws away exactly the information the interval is offering, which is that the data are entirely compatible with a real, positive difference too; there just isn't enough precision at n = 15 per group to rule zero out. A p-value collapses that whole interval down to one bit of information, pass or fail, and right at a boundary this close, that's precisely where the collapse costs you the most. I'd want a bigger sample before I'd rule out *any* effect of background music in general — this result argues for "Christmas specifically, at least as far as we can tell from here," not "definitely, only Christmas, case closed."

However, this type of post-hoc analysis has the potential for what is known as a *multiple comparisons problem*, which we might need to deal with.

Let's take a look **back in the slide deck...**

Remember, there are many other things that if we were doing ANOVA that we would also look to deal with - such as the various assumptions required of ANOVA, and so forth. But, they are beyond our scope in this class. Suffice to say that this has only scratched the surface of ANOVA so far.

## Reading the literature {#ANOVA-reading}

Experiments are common in some areas of management research, particularly in marketing, organisational behaviour and behavioural strategy, and ANOVA is how most of them are analysed. The simplest ones look just like the music study in this chapter: a few groups, one outcome, an F-test and some pairwise comparisons. But most published experiments go a step further and cross two manipulations in a *factorial design*, because the hypothesis is usually not "X affects Y" but "X affects Y, but only when Z." That's an *interaction*, and interactions are where most of the reading difficulty lies, whether they turn up in an ANOVA or a regression.

### Where you'll meet this

**The F-test.** Results are usually reported as something like F(2, 42) = 15.31, p < .001. The first number in brackets is the degrees of freedom for the groups (the number of groups minus one) and the second is for the error, which is roughly the total sample size minus the number of groups. So you can work backwards from the brackets to the sample size, which is worth doing when a paper is vague about how many people were in each condition. As in this chapter, a significant F only tells you that the groups aren't all the same. It doesn't tell you which ones differ, which is what the follow-up comparisons are for.

**Eta-squared and its relatives.** The effect size in this chapter was generalised eta-squared, the proportion of the variation in the outcome explained by group membership. With one factor, that's the same idea as R² in a regression. Papers also report plain eta-squared (η²) and, very often, *partial* eta-squared (η²p). In a one-factor design they're all the same number. In a factorial design they're not. Partial eta-squared leaves the other factors' effects out of the denominator, so it's usually larger, and partial eta-squareds from different designs can't be compared directly with each other [@lakens2013calculating]. You'll also see .01, .06 and .14 described as small, medium and large, and those labels deserve the same scepticism as Cohen's labels for d in Chapter \@ref(t-test).

**Factorial designs and interactions.** In a 2 × 2 design, say a price promotion (yes or no) crossed with brand familiarity (known or unknown brand), the analysis gives two *main effects* and one *interaction*. The main effect of the promotion is its average effect across both kinds of brand. The interaction asks whether the promotion's effect is different for known and unknown brands. When there's a real interaction, the main effects can be badly misleading: a promotion that helps unknown brands and does nothing for known ones has a positive "main effect" that describes neither. So when a paper reports an interaction, the main effects deserve much less attention than they usually get, and the *simple effects* (the effect of one factor at each level of the other) deserve much more. It also matters what shape the interaction takes. An effect that reverses direction is a very different finding from one that merely shrinks, and shrinking interactions in particular need far larger samples to detect reliably than most experiments have.

**Interactions in regression.** The same idea appears in survey papers as *moderation*: a regression with X, Z and their product X × Z. Two things commonly go wrong in reading these. First, once the product term is in the model, the coefficients on X and Z aren't main effects any more. The coefficient on X is the effect of X when Z equals zero, which may be a meaningless value, such as a firm with zero employees. That's one reason papers *centre* their variables (subtract the mean), which makes zero mean "average." Second, papers often claim that centring was done to reduce multicollinearity. It doesn't: centring changes how the lower-order coefficients are interpreted, but it leaves the precision of the estimates and the test of the interaction exactly as they were [@echambadi2007mean]. Look for a plot or table of *simple slopes*, which shows the effect of X at a few chosen values of Z. Without one, it's very hard to tell what an interaction actually means.

### The connection back

An ANOVA is a regression with the groups coded as dummy variables, just as the t-test in Chapter \@ref(t-test) was. The F-test asks whether the group means are further apart than the variation within the groups would lead you to expect, and eta-squared is R² by another name. What the factorial design adds is the question of whether one effect depends on another. That's usually the most interesting question in a paper, and the easiest one to misread.

### Questions worth asking

1. **Which effect size is reported?** Is it being compared with others calculated the same way?
2. **Is there an interaction?** If so, are the main effects being interpreted as if there weren't?
3. **What shape is the interaction?** Does an effect reverse, or just get smaller? Is there a plot of the simple effects or simple slopes?
4. **In a regression with a product term, what does zero mean for each variable?** What, therefore, does each coefficient actually describe?
5. **How many people were in each cell?** You can work it out from the degrees of freedom if the paper doesn't say.

### One red flag

A subtle interaction from a small experiment. A typical example is a 2 × 2 design with twenty or thirty people per cell, reporting that an effect is present under one condition and weaker under another. That's precisely the kind of finding that needs a large sample, and precisely the kind of design that doesn't have one. Put it next to what Chapter \@ref(Issues) says about power and about how many ways there are to analyse an experiment, and ask how much weight the finding can really bear. Then look at whether anyone has replicated it.

### If you want to read more

Julie Irwin and Gary McClelland's "Misleading heuristics and moderated multiple regression models" [@irwin2001misleading] is written for marketing researchers and goes through the most common mistakes in reading and reporting interactions, including the ones above. It's a little more technical than most of the readings I've suggested, but it's short, and it's the paper I'd give anyone about to test a moderation hypothesis.

A condensed version of these questions, combined with those from the other chapters and arranged in the order you'd meet them when reading a paper, is in Appendix \@ref(reader-guide).
