# Issues with Significance Testing {#Issues}

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(readxl)
library(ggplot2)
library(ggpubr)
library(mosaic)
library(dplyr)
library(rstatix)
library(pwr)

#library(waffle) NOTE: Waffle package removed from CRAN and not compatible with latest version of R. If it returns, go back to waffle code not hacky Base R code
```

This chapter provides examples and analysis for the third part of Lecture 8, and also parts of Lecture 9.

In this chapter, I'll explore examples of two common 'issues' or 'debates' regarding statistical significance testing, applied to two of our prior examples.

But first, let's unpack more about the Christmas music study, and the multiple comparisons problem.

## More on the Multiple Comparisons Problem

In this section, I'll provide some visualizations which can help us to understand the problem with multiple comparisons.

However, it's worth noting that the results depend on a number of assumptions about the research and what the real world actually looks like. The latter for sure we can never know. So, do bear that in mind.

First, let's visualize the situation I just discussed in the slides. We test 100 hypotheses, and in every case, the null is true in the population (remember, we cannot know the true state of the population in reality). Below, each grey box represents a hypothesis we test.

```{r}
#UGLY BASE R CODE BELOW. Note that this is not fully thought through and may contain unnecessary code. Code taken from https://stats.stackexchange.com/q/143424 and modified by me
waffle <- function(x, rows, cols = seq_along(x), ...) {
  xx <- rep(cols, times = x)
  lx <- length(xx)
  m <- matrix(nrow = rows, ncol = (lx %/% rows) + (lx %% rows != 0))
  m[1:length(xx)] <- xx

  op <- par(no.readonly = TRUE)
  on.exit(par(op))

  par(list(...))
  plot.new()
  o <- cbind(c(row(m)), c(col(m))) + 1
  plot.window(xlim = c(0, max(o[, 2]) + 1), ylim = c(0, max(o[, 1]) + 1),
              asp = 1, xaxs = 'i', yaxs = 'i')
  rect(o[, 2], o[, 1], o[, 2] + .85, o[, 1] + .85, col = c(m), border = NA)

  invisible(list(m = m, o = o))
}


cols <- c("lightgrey", "red", "#00BFC4", "#C77CFF")
m <- waffle(c(100, 0, 0, 0), rows = 10, cols = cols, mar = c(0,0,0,7),
            bg = 'white')
#legend('right', legend = LETTERS[1:2], pch = 15, col = cols, pt.cex = 2,
      # bty = 'n')

```

Now, remember, **even if all null hypotheses are true**, our significance level, or *p* = 0.05, so we have to expect that *on average* we would expect **5** of these hypotheses tests to return *p*-values \< 0.05, and thus we would reject the null. This is known as a **false positive**.

You might wonder how we could reduce the chances of false positive to zero. You could do this by setting your required *p* to be exactly zero. Do you see the problem here? You would certainly solve the false positive problem. But, you would also accept *all* null hypotheses. You would never actually find a statistically significant result in your research. Unless the null hypothesis really *was* true in the population in every single case, you would be inevitably missing out on making some discoveries.

Moving on, let's visualize the situation in hand here. Below, the same 100 hypothesis tests (remember, the null is true in all cases), but this time I have marked in red the expected level of false positives.

```{r}
#UGLY BASE R CODE BELOW. Note that this is not fully thought through and may contain unnecessary code. Code taken from https://stats.stackexchange.com/q/143424 and modified by me
waffle <- function(x, rows, cols = seq_along(x), ...) {
  xx <- rep(cols, times = x)
  lx <- length(xx)
  m <- matrix(nrow = rows, ncol = (lx %/% rows) + (lx %% rows != 0))
  m[1:length(xx)] <- xx

  op <- par(no.readonly = TRUE)
  on.exit(par(op))

  par(list(...))
  plot.new()
  o <- cbind(c(row(m)), c(col(m))) + 1
  plot.window(xlim = c(0, max(o[, 2]) + 1), ylim = c(0, max(o[, 1]) + 1),
              asp = 1, xaxs = 'i', yaxs = 'i')
  rect(o[, 2], o[, 1], o[, 2] + .85, o[, 1] + .85, col = c(m), border = NA)

  invisible(list(m = m, o = o))
}


cols <- c("lightgrey", "red", "#00BFC4", "#C77CFF")
m <- waffle(c(95, 5, 0, 0), rows = 10, cols = cols, mar = c(0,0,0,7),
            bg = 'white')
#legend('right', legend = LETTERS[1:2], pch = 15, col = cols, pt.cex = 2,
      # bty = 'n')

```

Now, that is just the *expected* level for every 100 tests. If you remember from the discussion in Section \@ref(conf), it may not be you get *exactly* 5 false positives in 100 tests, just that you would expect the long run average number of false positives to be 5%, in an all-null world.

So, we can look at it through a different lens. What is the probability that you would get **at least one false positive** in any given batch of multiple tests.

That is known as the *familywise error rate*, and it applies whenever you can define a cluster or 'family' of tests.

It's the same principle as rolling a 6-sided die, and needing to score a '6'. Think of the '6' as the 'false positive' here.

In a single die role, the probability of getting a '6' is $\frac{1}{6}$, and of course the equivalent probability of *not* getting a '6' is $1-\frac{1}{6}=\frac{5}{6}$.

*But*, what if you get to roll the die 3 times? It's fairly easy to intuit that, even though the probability of getting a '6' on any single roll stays the same, the probability of getting a single '6' in 3 rolls is higher. In fact, we can calculate it.

To do so, we use the probability of *not* getting a 6, as follows:

*P*(no 6)=$(\frac{5}{6})^{3} = 0.579$

Recalling the laws of probability then...

*P*(at least one 6) = 1 - *P*(no 6) = 1 - .579 = .421

So, *now* we have a 42% probability of at least one 6 in 3 rolls, even though the probability of getting a 6 in any single roll is unchanged.

We can calculate the familywise error rate for any set of multiple comparisons in exactly the same way.

One of the key advantages of ANOVA is that it is a single test, and thus does not fall prey to the multiple comparisons problem. However, in the case of *post-hoc* testing for the Christmas music study (Study 2), we are doing 3 pairwise comparisons (i.e. 3 tests) across the 3 groups, with a stated significance level of 0.05.

For 1 test, the false positive rate is $1 - 0.95 = .05$

But, for 3 tests, the false positive rate is $1-0.95^{3}=0.14$

We are 'rolling the dice' multiple times.

In fact, the familywise error rate is easy to see for yourself, once you know the formula: it's simply $1-(1-\alpha)^n$, where $\alpha$ is your chosen significance level and *n* is the number of tests you're running. The figure below plots that out for a few different significance levels, so you can see how quickly the error rate climbs as the number of comparisons grows.

I first saw this demonstrated properly using a lovely interactive calculator built by Dr Daniel Roelfs, during his time at the Norwegian Centre for Mental Disorders Research (NORMENT) in Oslo, and he kindly gave permission for me to reproduce the idea here. I still use his live version in the lecture itself, where you can drag the sliders around and watch the number update in real time - genuinely a nicer way to build intuition for this than a static plot. Check out his website: <https://danielroelfs.com/about/>. A book page can't rely on an external app staying up forever for every reader, though, so the version below reproduces the same idea as a static figure instead.

```{r}
#Familywise error rate = 1 - (1 - alpha)^n, plotted across a range of test counts
#reproduces the idea behind Dr Daniel Roelfs' interactive FWER calculator, see text above

fwer_df <- expand.grid(n = 1:30, alpha = c(0.05, 0.01, 0.001))
fwer_df$FWER <- 1 - (1 - fwer_df$alpha) ^ fwer_df$n
fwer_df$alpha <- factor(fwer_df$alpha,
                         levels = c(0.05, 0.01, 0.001),
                         labels = c("alpha = 0.05", "alpha = 0.01", "alpha = 0.001"))

#the two values Nick reads off the live app in the lecture: 3 comparisons (the post-hoc
#case just above) and 26 comparisons (the football goals case a bit further below)
fwer_points <- data.frame(
  n = c(3, 26),
  FWER = c(1 - 0.95^3, 1 - 0.95^26),
  label = c("3 comparisons\nFWER = 0.14\n(the post-hoc case)",
            "26 comparisons\nFWER = 0.74\n(the football goals case)"),
  hjust = c(0, 1),
  nudge_x = c(1.5, -1.5)
)

ggplot(fwer_df, aes(x = n, y = FWER, colour = alpha)) +
  geom_line(linewidth = 1) +
  geom_point(data = fwer_points, aes(x = n, y = FWER), colour = "black", size = 2, inherit.aes = FALSE) +
  geom_label(data = fwer_points,
             aes(x = n + nudge_x, y = FWER + 0.05, label = label, hjust = hjust),
             colour = "black", fill = "white", size = 3, inherit.aes = FALSE) +
  scale_x_continuous(limits = c(0, 30)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Number of comparisons", y = "Familywise error rate",
       colour = "Significance level",
       title = "Familywise Error Rate vs Number of Comparisons") +
  theme_bw()
```

Watching that curve climb feels kind of scary, right? There are many ways to deal with the problem, but they pretty much all amount to making some correction to the required significance level, making it more stringent, in order to reduce the false positives.

For example, in the Christmas-music post-hoc tests in Chapter \@ref(ANOVA), we used Tukey's test, which uses a correction for multiple comparisons. Another method dealt with below is the *Bonferoni Correction*, which is generally the most well-known of them.

## The Bonferroni Correction: Rate of Change in Football Goals per Season

Here, I will expand on the football goals rate-of-change example from Chapter \@ref(statistics), to incorporate corrections for the multiple comparisons problem. Again, I'm using data from <https://www.footballhistory.org/league/premier-league-statistics.html>

I'm building on the z-test table from the end of Chapter \@ref(statistics). But here, using the z-scores in the data file, I have calculated the p-value (2 tailed as we do not hypothesize a direction for the difference) for the z-scores for the difference between each season, year-on-year.

```{r}
EPLP<-read_excel("Data/EPLGOALSP.xlsx")
summary(EPLP)
head(EPLP)
```

Here, we can plot the p-values (2-tailed), and again we see that two seasons seem to have significant differences. In other words, the p-values are less than 0.05 for the test as to whether the number of goals scored differs from the season before

```{r}
#visualize the z values simply with control lines
U <- 0.05
p <- ggplot(EPLP, aes(x=Season, y=p2)) + geom_point() +theme(axis.text.x = element_text(angle = 90))+scale_x_discrete(name="Country")
p <- p+ geom_hline(aes(yintercept=U))
p

```

We can see that the 1999-2000 season, and the 2009-10 seasons have p values less than 0.05

The question is **are we suffering from the multiple comparisons problem**? Should we correct for it?

It's hard to say actually. Of course, we are indeed running multiple tests, 26 in fact. So, the chance of a false positive is high. The Bonferroni correction would immediately reduce the false positive chances, but at what cost?

Let's see how this would work. In order to calculate a Bonferroni correction, you can either adjust the p-value directly that you calculate for each test, or instead simply adjust the 'cutoff' value for p, known as the *critical p*, to lower it from 0.05 and make the significance test 'harder' to pass. The formula to adjust the cutoff value is simply:

$\alpha_b = (\frac{\alpha}{n})$

Where

$\alpha_b$ = Bonferroni-adjusted critical p value

$\alpha$ = original critical p value (here this is 0.05)

*n* = number of comparisons (here this is 26)

So, the formula gives us a new Bonferroni-adjusted critical p value of 0.0019

Let's see what happens with this new critical p value:

```{r}
#visualize the z values simply with control lines
U <- 0.0019
p <- ggplot(EPLP, aes(x=Season, y=p2)) + geom_point() +theme(axis.text.x = element_text(angle = 90))+scale_x_discrete(name="Country")
p <- p+ geom_hline(aes(yintercept=U))
p

```

We can see that none of our tests now rejects the null. The Bonferroni correction is known as a highly conservative test. That is, it is based on the idea that the *null hypothesis is true* in each case in the population. We can thus consider it as the most stringent and conservative way to correct for the chance of false positives when doing multiple comparisons.

But, as I suggested above, that might not always be the best idea. Indeed, if you want to totally avoid any chance of a false positive, why not simply make the required alpha 0? Then, you would never get a false positive. Of course, you would never detect a **true positive** either.

What if it is the alternative hypothesis (that is, the H of an effect existing) that is true in all cases? In such cases, there can of course be *no false positives*. Therefore, in such a situation you would be *increasing the chances of a false negative* by reducing the chances of a false positive. So, what are the potential costs of each of these mistakes?

For example, Thomas Perneger's 1998 paper in the BMJ is scathing about the Bonferroni adjustment. Take a look at <https://www.bmj.com/content/316/7139/1236.full>

Mind you, I am not saying that's the final word, just that there are multiple perspectives on the issues!

It's never as simple as it seems when making statistical decisions, is it?

## Statistical Power: Brief Demonstration

Different types of analysis and research design require different types of power calculation, so it is hard to give a uniform example. But, for simplicity's sake, let's calculate the required sample size for the Christmas music study (Study 2) we conducted earlier, in Chapter \@ref(ANOVA).

Remember, really, we should have done this **before collecting data**.

To calculate power, all we need the parameters of the experiment and analysis design. As such, it is easily possible to do this before collecting data, and to design your studies around it. Really, we should do this a lot more in business and management - it's routine in fields like medicine.

So, we had 3 groups, and used ANOVA

Let's set a significance of 0.05, a required power of 0.8, and assume the effect size is moderate (say 0.25)

```{r}
pwr.anova.test(k=3,f=.25,sig.level=.05,power=.8)
```

So, we really wanted to have around 50 in each group to have an 80% chance of detecting a moderate effect presuming the null was true.

You can see that my study (with only 15 in each group) was rather underpowered — assuming a moderate effect. But now here's a slightly awkward wrinkle. If instead I take the effect size *this actual experiment* produced — a generalized eta-squared of 0.41 back in Chapter \@ref(ANOVA), which converts to a Cohen's *f* of about 0.84 — and plug that into the same calculation:

```{r}
pwr.anova.test(k=3,f=.84,sig.level=.05,power=.8)
```

That says I'd only have needed about 6 people per group for 80% power. That's *fewer* than the 15 I actually used, not closer to it from above — a different, and slightly uncomfortable, lesson from the one I set out to teach.

Here's the honest reading. What I've just done — taking the effect size my own small sample happened to produce, and treating it as the 'true' population effect size for a power calculation — is a well-known bad practice, sometimes called *observed* or *post-hoc* power. Small samples give you very noisy estimates of the true effect size: a difference that looks huge in a sample of 15 per group could easily be a real, more modest effect combined with a slice of good luck in exactly this data. So this isn't really evidence that a study of 15 would reliably detect an effect this size in general — it's evidence that *this specific sample* turned up a strikingly large difference. The honest way to plan a study is to commit to a plausible effect size **before** you collect any data (which is what the moderate f = .25 assumption above was trying to do, arbitrary as it was), not to reverse-engineer one from a result you already have. If anything, this is itself a small demonstration of why "just check whether the effect size you found was big enough" isn't a fix for a badly-powered study after the fact — the number you'd get depends entirely on which random sample you happened to draw, small samples especially.
