---
title: "Factor Analysis Setup"
output: html_document
date: "2022-10-26"
---


# Factor Analysis in R: Introduction

Set up for doing factor analysis

first install and load required packages

```{r}
install.packages("psych")
install.packages("psychTools")
install.packages("GPArotation")
install.packages("semTools")
```


```{r}
library(psych)
library(psychTools)
library(readxl)
library(GPArotation)
library(semTools)
```

Load data

You'll need to download the 'EXCERCISE_1_OC.xlsx' file from the OSF repository, and place it in a location of your choice, so you can point R to it, as I do below.

Remember, you'll need your code to point to the file location on your computer not mine! I keep all my data files in a single directory on my D drive. You need to work our your own setup.

```{r}
OC<-read_excel("D:/Dropbox/R_Files/Data/EXCERCISE_1_OC.xlsx")
```

On Friday, we will download the rest of this script, and you can follow along with it if you want to.


